Non-analogues in paleoecological reconstruction

### 2015-09-17 Paper_Markdown.Rmd
Lots of new text in the introduction.  Moved many plots into their own files (in R/figures).  Added some methods text.
New references.

Cut libraries that were unused.

Added a 'null' model.  Just randomly samples from the data without respect to underlying data.

Although we cannot know the values for temperature or precipitation during this time, statistical non-analogues do exist in a modern context, and we can construct false non-analogues using a modified h-block sampling procedure [*sensu*, @telford2009evaluation] that examines analogue distance rather than geographical distance.

@@

map <- map_data('world')

map <- subset(map, map$long > -160 & map$long < -30)
map <- subset(map, map$lat > 20 & map$lat < 83)

ggplot(data = data.frame(map), aes(long, lat)) + 
  geom_polygon(aes(group=group), color='blue', alpha=0.7) +
  geom_point(data = data.frame(climate), aes(x = LONDD, y = LATDD), alpha = 0.5) +
  theme_bw() +
  xlab('') + ylab('') +
  theme(axis.text=element_text(size=12, family='serif'),
        panel.background = element_rect(fill = 'lightblue')) +
  coord_map(projection='albers', lat0=35, lat1=55, xlim=c(-100, -60), ylim=c(35, 80))

@@

WAPLS does this and that. . .

BRT [@de2007boosted] and randomForest [@breiman2001random] are ensemble machine-learning approaches which have seen increasing use in fossil pollen literature during the past few years [][SOME REFS HERE IF NECESSARY ... GORING ET AL. 2010, ZHAO ET AL. 2012, SALONEN ET AL. 2014...]. Both methods are based on creating an ensemble of regression tree models with all pollen taxa used as predictors and an environmental variable of choice as the response. The randomForest creates an ensemble of independent tree models (here 500 trees), while using bootstrap samples of the calibration data when building each tree, and randomly selected subsets of the predictor set when determining each tree split, thus creating diversity in the tree ensemble. In BRT, by comparison, the tree ensemble is built sequentially, with each added tree aiming to explain the residuals of the previously fitted ensemble. The final BRT model can thus be likened to an additive regression model in which the individual terms are regression trees.

>>>>

Although we cannot know the values for temperature or precipitation during this time, statistical non-analogues do exist in a modern context, and we can construct false non-analogues using a modified h-block sampling procedure [*sensu*, @telford2009evaluation] that examines analogue distance rather than geographical distance.  The result of no-analogue prediction is likely to vary with the method used for climate reconstruction, and possibly the climate variable (or distribution of the climate variable) of interest. This has been shown to occur with reconstructions from mixed depositional environments [@goring2010pollen], and more broadly, related families of methods are likely to behave similarly [KNUTTIPaper, NEWCANNONPAPER].

The Modern Analogue Technique is effectively a k-nearest neighbors method, generally using a squared-chord metric of distance.  MAT is used commonly in paleoecological reconstruction [@overpeck1985quantitative], but the k-nearest neighbor alogrithm is common to big data analysis. With MAT, the effects of moving toward no-analogue space will likely vary depending on the location of the sample in multi-dimentsional space.  We expect that a sample close to the average of all samples will show increasing predictive variability, but little effect on the mean prediction as the closest analogues become progressvely further from the focal sample, but the average of distant samples remains similar. In general, with increasing analogue distance we expect samples should revert toward the mean, while also showing higher variability.  Of all the methods, MAT's dependence on the local neighborhood likely suggests that we should expect a higher increase in variance with increasing analogue distance.

When WA is used for reconstruction in no-analogue space it is likely that species combinations found along the climatic gradient of interest are absent, meaning higher variability of prediction and greater likelihood that species optima or tolerances will be misinterpreted.  This is likely to be most significant for taxa with bimodal distributions,

@@

source('R/figures/figure1_map.R')

map_plot

@@

WAPLS is a modification of WA, whereby the residuals for each component are again weighted and regressed and the number of components are chosen by minimizing the root mean squared error (RMSE) of the model [@ter1993weighted].

BRT [@de2007boosted] and randomForest [@breiman2001random] are ensemble machine-learning approaches which have seen increasing use in fossil pollen literature during the past few years [@goring2010pollen;@salonen2014reconstructing;@salonen2012north;@zhao2012application]. Both machine learning methods are based on creating an ensemble of regression tree models with all pollen taxa used as predictors and an environmental variable of choice as the response. The random forest creates an ensemble of independent tree models (here 500 trees), while using bootstrap samples of the calibration data when building each tree, and randomly selected subsets of the predictor set when determining each tree split, thus creating diversity in the tree ensemble. In BRT, by comparison, the tree ensemble is built sequentially, with each added tree aiming to explain the residuals of the previously fitted ensemble. The final BRT model can thus be likened to an additive regression model in which the individual terms are regression trees.
